\begin{center}
{\Large \textbf{Saturn: Zero-Configuration AI Service Discovery\\via mDNS and DNS-SD}}

\vspace{1em}
Joey Pereira
\end{center}

\vspace{1em}

Generative AI services gate access behind per-user subscriptions, API keys, and manual endpoint configuration---barriers that scale with the number of users and applications on a network.
Saturn provisions AI the way printers and speakers already provision themselves: through Multicast DNS (mDNS) and DNS-based Service Discovery (DNS-SD), protocols that ship on every major operating system.
An administrator registers AI endpoints under the service type \texttt{\_saturn.\_tcp.local.}; every device on the network segment discovers them without accounts, credentials, or configuration files.
Seven implementations across four languages and four independent mDNS libraries interoperate on this service type with no shared discovery code, demonstrating that the protocol specification---not a reference implementation---carries the design.
A cognitive walkthrough across three user personas (administrator, developer, end user) measures a 53\% reduction in total configuration steps compared to traditional per-user API key distribution, with developer effort dropping 79\% and end-user payment friction eliminated entirely.
Cloud-facing deployments rotate ephemeral API keys every five minutes; a captured credential expires before an attacker leaves the building.
Saturn's broadcast model inherits known mDNS privacy risks---passive enumeration, device-name leakage---and enterprise WiFi access-point isolation blocks multicast entirely, confining viable deployment to home, office, and lab networks.
These trade-offs are documented in existing literature and addressable without destroying the zero-configuration property.
