\chapter{Conclusion}
\label{ch:conclusion}

Saturn demonstrates that AI services can be provisioned like printers, DHCP addresses, and file shares---announced on the local network and discovered without user configuration. The protocol required no changes to mDNS/DNS-SD; the contribution is evidence that existing zero-configuration infrastructure carries AI-specific metadata without adaptation.

Independent implementations interoperate on a shared service type because the protocol specification, not any reference implementation, defines conformance (\argref{V1}). Network-level provisioning shifts the cost structure: one administrator configures a beacon, and every user on the segment discovers services without individual setup (\argref{V2}). The security model trades broadcast exposure for temporal containment---ephemeral keys expire before leaked credentials become useful, at the cost of local-segment visibility that static keys avoid (\argref{V3}).

The structural insight is the asymmetry. Per-user provisioning scales linearly: each person added to a network repeats the same credential acquisition, endpoint configuration, and billing setup. Saturn makes provisioning a per-network constant---one daemon, configured once, serving every device that joins the segment.

Two constraints bound these findings. AP isolation blocks multicast between wireless clients, and the institutions where shared AI provisioning would help most---universities, public libraries, community colleges---are the institutions most likely to run that configuration (\argref{A17}). This is the cost of the design's central choice: zero-configuration requires multicast, and multicast requires cooperative infrastructure. Second, all implementations originate from one developer; independent adoption would strengthen the interoperability evidence.

Wide-area DNS-SD and relay architectures can extend discovery across subnets, but each reintroduces centralized authority. The open question is where that boundary falls---how much coordination overhead can return before the zero-configuration property ceases to hold.

The protocols exist and the implementations interoperate. The remaining barriers are both technical---AP isolation, schema drift, single-segment scope---and organizational---convincing network administrators that AI belongs alongside the services they already provision. Saturn establishes that the technical path is open.
